### 2018-07-07 paper/2-background.tex
Final proofread

Build systems automate the execution of repeatable tasks for individual users
and large organisations. In this section we explore the design space of build
systems, using four concrete examples: \Make~\cite{feldman1979make},
\Shake~\cite{mitchell2012shake}, \Bazel~\cite{bazel}, and
\Excel~\cite{advanced_excel}.
We have carefully chosen these four to illustrate the various axes on which
build systems differ; we discuss many other notable examples of build systems,
and their relationships, in~\S\ref{sec-related}.

@@

\noindent
The above makefile lists three tasks: (i) compile a utility library comprising
files \cmd{util.h} and \cmd{util.c} into \cmd{util.o} by
executing\footnote{In this example we pretend \cmd{gcc} is a pure function for
the sake of simplicity. In reality, there are multiple versions of \cmd{gcc}
and the actual binary used to compile and link files, and standard libraries
such that \cmd{stdio.h}, are often also listed as dependencies.} the command
\cmd{gcc -c util.c}, (ii) compile the main source file \cmd{main.c} into
\cmd{main.o}, and (iii) link object files \cmd{util.o} and \cmd{main.o} into the
executable \cmd{main.exe}. The makefile contains the complete information about
the \emph{task dependency graph}, which is shown in Fig.~\ref{fig-make}(a).

>>>>

Build systems automate the execution of repeatable tasks for individual users
and large organisations. In this section we explore the design space of build
systems, using four concrete examples: \Make~\cite{feldman1979make},
\Shake~\cite{mitchell2012shake}, \Bazel~\cite{bazel}, and
\Excel~\cite{advanced_excel}.
We have carefully chosen these four to illustrate the various axes on which
build systems differ; we discuss many other notable examples of build systems,
and their relationships, in~\S\ref{sec-implementations} and~\S\ref{sec-related}.

@@

\noindent
The above makefile lists three tasks: (i) compile a utility library comprising
files \cmd{util.h} and \cmd{util.c} into \cmd{util.o} by
executing\footnote{In this example we pretend \cmd{gcc} is a pure function for
the sake of simplicity. In reality, there are multiple versions of \cmd{gcc};
the actual binary used to compile and link files, and standard libraries
such that \cmd{stdio.h}, are often also listed as dependencies.} the command
\cmd{gcc -c util.c}, (ii) compile the main source file \cmd{main.c} into
\cmd{main.o}, and (iii) link object files \cmd{util.o} and \cmd{main.o} into the
executable \cmd{main.exe}. The makefile contains the complete information about
the \emph{task dependency graph}, which is shown in Fig.~\ref{fig-make}(a).

### 2018-07-07 paper/3-abstractions.tex
Final proofread

\noindent
Here \hs{Const} is a standard Haskell type defined in Fig.~\ref{fig-stdlib}. We
instantiate \hs{f} to \hs{Const}~\hs{[@@k]}. So a value of type \hs{f}~\hs{v},
or in this case \hs{Const}~\hs{[@@k]}~\hs{v}, contains no value \hs{v}, but does
contain a list of keys of type \hs{[@@k]} which we use to record dependencies.
The \hs{fetch} callback that we pass to \hs{task} records a single dependency;
and the standard definition of \hs{Applicative} for \hs{Const} (which we give
in Fig.~\ref{fig-stdlib}) combines the dependencies from different parts of the
task. Running the task with \hs{f}~=~\hs{Const}~\hs{[@@k]} will thus
accumulate a list of the task's dependencies -- and that is what
\hs{dependencies} does:
\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
@\ghci@ dependencies $ fromJust $ sprsh1 "A1"
[@@]
\end{minted}
\begin{minted}[xleftmargin=10pt]{haskell}
@\ghci@ dependencies $ fromJust $ sprsh1 "B1"
["A1", "A2"]
\end{minted}
\vspace{1mm}

>>>>

\noindent
Here \hs{Const} is a standard Haskell type defined in Fig.~\ref{fig-stdlib}. We
instantiate \hs{f} to \hs{Const}~\hs{[@@k]}. So a value of type \hs{f}~\hs{v},
or in this case \hs{Const}~\hs{[@@k]}~\hs{v}, contains no value \hs{v}, but does
contain a list of keys of type \hs{[@@k]} which we use to record dependencies.
The \hs{fetch} callback that we pass to \hs{task} records a single dependency;
and the standard definition of \hs{Applicative} for \hs{Const} (which we give
in Fig.~\ref{fig-stdlib}) combines the dependencies from different parts of the
task. Running the task with \hs{f}~=~\hs{Const}~\hs{[@@k]} will thus
accumulate a list of the task's dependencies -- and that is what
\hs{dependencies} does:
\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
@\ghci@ dependencies $ fromJust $ sprsh1 "B1"
["A1","A2"]
\end{minted}
\begin{minted}[xleftmargin=10pt]{haskell}
@\ghci@ dependencies $ fromJust $ sprsh1 "B2"
["B1"]
\end{minted}
\vspace{1mm}

### 2018-07-07 paper/4-build.tex
Explain the Frankenbuild example
Mostly whitespace tweaks
Final proofread

To reduce the number of aborts (often to zero) Excel records the
discovered task order in its \emph{calc chain}, and uses it as the
starting point for the next build (\S\ref{sec-background-excel}).
\Bazel's restarting scheduler does not store the discovered order
between build runs; instead, it stores the most recent task dependency
information. Since this information may become outdated, \Bazel may
also need to abort a task if a newly discovered dependency is
out-of-date.

@@

Suppose the scheduler decides that a key should be brought up to date. The
next question is: does any work need to be done, or is it already up to date?
Or, in a cloud-build system, do we have a cached copy of the value we need?
These questions can be addressed in one of four fundamental ways, with a number
of tweaks and variations within them.

@@

\Excel models the dirty bit approach most directly, having an actual dirty bit
associated with each cell, marking the cell dirty if the user modifies it.
It also marks dirty all cells that (transitively) depend on the modified cell.
Excel does not record dynamic dependencies of each cell; instead it computes a
\emph{static over-approximation} -- it is safe for it to make more cells dirty
than necessary, but not vice versa. The over-approximation is as follows: a cell
is marked dirty if its formula statically refers to a dirty cell, or if the
formula calls a function like \cmd{INDIRECT} whose dependencies cannot be
guessed from the formula alone. The over-approximation is clear for
\cmd{INDIRECT}, but it is also present for \cmd{IF}, where both branches are
followed even though dynamically only one is used.

@@

\noindent
The function \hs{recordCT} looks like \hs{recordVT}, but instead of just passing
the hash of the resulting value, we require the actual value. The \hs{verifyVT}
has been replaced with \hs{constructCT}, which instead of taking the hash of the
current value as \emph{input}, returns a list of suitable values as \emph{output}.
If the current value in the store
matches one of the possible values, the build can skip this rule. If the
resulting list is empty, the key must be rebuilt. However, if the current value
does not match the store, and there is a possible value, we can use any value
from the constructive list \emph{without} doing any work to build it, and copy
it into the store.

@@

Constructive traces always verify keys by looking at their immediate
dependencies, which must have first been brought up to date, meaning that the
time to verify a key is based on the number of transitive dependencies. A
\emph{deep} constructive trace optimises this process by only looking at the
terminal \emph{input keys}, ignoring any intermediate dependencies. The operations
capturing this approach are the same as for constructive traces
in~\S\ref{sec-constructive-traces}, but we use the names \hs{recordDCT} and
\hs{constructDCT}, where the underlying \hs{DCT} representation need only record
information about hashes of inputs, not intermediate dependencies.

@@

% Concretely, if
% the result of building a key given a set of immediate dependencies can vary,
% then it is possible that key \hs{k} will have been computed locally, but
% something \emph{using} \hs{k} will have been taken as a constructive input using
% a different value for \hs{k}. If another rule combines those two values then two
% different values of \hs{k} will have been used in one rule, leading to problems
% such as link-errors for typical compilers -- a phenomenon known as a
% \emph{Frankenbuild}~\cite{esfahani2016cloudbuild}.

@@

\begin{enumerate}
\item Any system using verifying traces is unlikely to see significant benefit
from storing more than one \hs{Trace} per key\footnote{There is a small chance
of a benefit if the dependencies change but the result does not, and then the
dependencies change back to what they were before.}. Therefore, such systems can
store \hs{Map}~\hs{k}~\hs{(Trace}~\hs{k}~\hs{v}~\hs{(Hash}~\hs{v))}, where the
initial \hs{k} is the \hs{key} field of \hs{Trace}.

@@

\item Many cloud build systems store hashes of values in the trace information,
then have a separate content-addressable cache which associates hashes with
their actual contents.
\end{enumerate}

@@

\begin{table}
\vspace{-1mm}
\smaller
\centering
\begin{tabular}{l||c|c|c}
\hline
 & \multicolumn{3}{c}{Scheduling strategy} \\
Out-of-date key strategy  & Topological\hspace{2mm}\S\ref{sec-topological} & Restarting\hspace{2mm}\S\ref{sec-restarting} & Suspending\hspace{2mm}\S\ref{sec-suspending}    \\\hline
\hline
Dirty bit\hfill\S\ref{sec-dirty-bit}                                                             & \Make       & \Excel & -              \\\hline
Verifying traces\hfill\S\ref{sec-verifying-traces}                                               & \Ninja      & -      & \Shake         \\\hline
Constructive traces\hspace{2mm}\hfill\S\ref{sec-constructive-traces}                             & \CloudBuild & \Bazel & -              \\\hline
Deep constructive traces\hspace{2mm}\hfill\S\ref{sec-deep-constructive-traces} & \Buck       & -      & \Nix           \\\hline
\end{tabular}
\vspace{1mm}
\caption{Build systems \`a la carte.\label{tab-build-systems}}
\vspace{-8mm}
\end{table}

With the information in this section we can build Table~\ref{tab-build-systems},
which tabulates combinations of the scheduling strategy and the out-of-date key
strategy, providing 12~possible build systems, 8~of which are inhabited
by existing build systems (we discuss these systems in~\S\ref{sec-background} and
\S\ref{sec-related-build}). Of the remaining 4~spots, all result in workable
build systems. The most interesting unfilled spot in the table is suspending
constructive traces, which would provide many benefits, and which we title
\Cloud \Shake and explore further in~\S\ref{sec-implementation-cloud}.
% (as we plan on extending \Shake to occupy that spot)

>>>>

To reduce the number of aborts (often to zero) \Excel records the
discovered task order in its \emph{calc chain}, and uses it as the
starting point for the next build (\S\ref{sec-background-excel}).
\Bazel's restarting scheduler does not store the discovered order
between build runs; instead, it stores the most recent task dependency
information. Since this information may become outdated, \Bazel may
also need to abort a task if a newly discovered dependency is out of date.

@@

Suppose the scheduler decides that a key should be brought up to date. The next
question is: does any work need to be done, or is the key already up to date?
Or, in a cloud-build system, do we have a cached copy of the value we need?
These questions can be addressed in one of four fundamental ways, with a number
of tweaks and variations within them.

@@

\Excel models the dirty bit approach most directly, having an actual dirty bit
associated with each cell, marking the cell dirty if the user modifies it.
It also marks dirty all cells that (transitively) depend on the modified cell.
\Excel does not record dynamic dependencies of each cell; instead it computes a
\emph{static over-approximation} -- it is safe for it to make more cells dirty
than necessary, but not vice versa. The over-approximation is as follows: a cell
is marked dirty if its formula statically refers to a dirty cell, or if the
formula calls a function like \cmd{INDIRECT} whose dependencies cannot be
guessed from the formula alone. The over-approximation is clear for
\cmd{INDIRECT}, but it is also present for \cmd{IF}, where both branches are
followed even though dynamically only one is used.

@@

\noindent
The function \hs{recordCT} looks like \hs{recordVT}, but instead of just passing
the hash of the resulting value, we require the actual value. The \hs{verifyVT}
has been replaced with \hs{constructCT}, which instead of taking the hash of the
current value as \emph{input}, returns a list of suitable values as \emph{output}.
If the current value in the store matches one of the possible values, the build
can skip this key. If the resulting list is empty, the key must be rebuilt.
However, if the current value does not match the store, and there is a possible
value, we can use any value from the constructive list \emph{without} doing any
work to build it, and copy it into the store.

@@

Constructive traces always verify keys by looking at their immediate
dependencies, which must have first been brought up to date, meaning that the
time to verify a key depends on the number of transitive dependencies. A
\emph{deep} constructive trace optimises this process by only looking at the
terminal \emph{input keys}, ignoring any intermediate dependencies. The operations
capturing this approach are the same as for constructive traces
in~\S\ref{sec-constructive-traces}, but we use the names \hs{recordDCT} and
\hs{constructDCT}, where the underlying \hs{DCT} representation need only record
information about hashes of inputs, not intermediate dependencies.

@@

@@

\begin{itemize}
\item Any system using verifying traces is unlikely to see significant benefit
from storing more than one \hs{Trace} per key\footnote{There is a small chance
of a benefit if the dependencies change but the result does not, and then the
dependencies change back to what they were before.}. Therefore, such systems can
store \hs{Map}~\hs{k}~\hs{(Trace}~\hs{k}~\hs{v}~\hs{(Hash}~\hs{v))}, where the
initial \hs{k} is the \hs{key} field of \hs{Trace}.

@@

\item Many cloud build systems store hashes of values in the trace information,
then have a separate content-addressable cache which associates hashes with
their actual contents.
\vspace{-1mm}
\end{itemize}

@@

\begin{table}
\vspace{-1mm}
\smaller
\centering
\begin{tabular}{l||c|c|c}
\hline
 & \multicolumn{3}{c}{Scheduling algorithm} \\
Rebuilding strategy  & Topological\hspace{2mm}\S\ref{sec-topological} & Restarting\hspace{2mm}\S\ref{sec-restarting} & Suspending\hspace{2mm}\S\ref{sec-suspending}    \\\hline
\hline
Dirty bit\hfill\S\ref{sec-dirty-bit}                                                             & \Make       & \Excel & -              \\\hline
Verifying traces\hfill\S\ref{sec-verifying-traces}                                               & \Ninja      & -      & \Shake         \\\hline
Constructive traces\hspace{2mm}\hfill\S\ref{sec-constructive-traces}                             & \CloudBuild & \Bazel & -              \\\hline
Deep constructive traces\hspace{2mm}\hfill\S\ref{sec-deep-constructive-traces} & \Buck       & -      & \Nix           \\\hline
\end{tabular}
\vspace{1mm}
\caption{Build systems \`a la carte.\label{tab-build-systems}}
\vspace{-8mm}
\end{table}

With the information in this section we can build Table~\ref{tab-build-systems},
which tabulates combinations of the scheduling algorithm and the rebuilding
strategy, providing 12~possible build systems, 8~of which are inhabited
by existing build systems (we discuss these systems in~\S\ref{sec-background} and
\S\ref{sec-related-build}). Of the remaining 4~spots, all result in workable
build systems. The most interesting unfilled spot in the table is suspending
constructive traces, which would provide many benefits, and which we title
\Cloud \Shake and explore further in~\S\ref{sec-implementation-cloud}.
% (as we plan on extending \Shake to occupy that spot)

### 2018-07-07 paper/5-implementations.tex
Final proofread

\noindent
A \hs{Scheduler} is a function that takes a \hs{Rebuilder} and uses
it to construct a \hs{Build} system, by choosing which keys to rebuild in which
order. The \hs{Rebuilder} makes uses of the persistent build information
\hs{ir}, while the scheduler might augment that with further persistent
information of its own, yielding \hs{i}.

@@

The scheduler calls the rebuilder, but passes it a \hs{fetch} function
that that latter calls when it needs the value of a dependent key.  This
callback returns control to the scheduler, which may in turn call the
rebuilder to bring the dependent key up to date, and so on.

@@

The rebuilder receives three arguments: a \hs{key}, its current \hs{value}, and
the applicative \hs{task} that can be used to rebuild the \hs{key} if necessary.
The rebuilder first decides if the \hs{key} is \hs{dirty} by consulting
\hs{modTimes}: if the \hs{key} is not found, that must mean it has never been
built before; otherwise \hs{modTimeRebuilder} can see if any of the \hs{task}'s
dependencies (computed by \hs{dependencies}) are out of date. If the \hs{key} is
\hs{dirty}, we use \hs{run}~\hs{task} to rebuild it, and update the state with
the new modification time of the \hs{key}\footnote{The real \Make
relies on the file system to track modification times, but we prefer
to make it explicit here.}; otherwise we can just return the current \hs{value}.

@@

The external system, which invokes \Excel's build engine, is required
to provide a transitively closed set of dirty bits.  That is, if a
cell is changed, its dirty bit is set, and the dirty bit of any cell
whose value might perhaps change as a result.  It is OK to mark too many
cells as dirty; but not OK to mark too few.

@@

Fig.~\ref{fig-cloud-implementations} shows our models of several cloud build
systems. \Bazel, \CloudBuild and \Cloud \Shake are based on constructive traces
(\S\ref{sec-constructive-traces}), whereas \Buck and \Nix use their deep
version (\S\ref{sec-deep-constructive-traces}).

@@

By switching to the \hs{topological} scheduler, we obtain a model of
Microsoft's \CloudBuild~-- an applicative build system that combines
conventional scheduling of statically known directed acyclic graphs of
dependencies with constructive traces~\cite{esfahani2016cloudbuild}. Note that
we need to convert a monadic \hs{ctRebuilder} into an applicative one by
applying an adapter function \hs{adaptRebuilder}, which unwraps a given
\hs{Task Applicative} and wraps it into \hs{Task Monad}.

>>>>

\noindent
A \hs{Scheduler} is a function that takes a \hs{Rebuilder} and uses
it to construct a \hs{Build} system, by choosing which keys to rebuild in which
order. The \hs{Rebuilder} makes use of the persistent build information
\hs{ir}, while the scheduler might augment that with further persistent
information of its own, yielding \hs{i}.

@@

The scheduler calls the rebuilder, but passes it a \hs{fetch} function
that the latter calls when it needs the value of a dependent key.  This
callback returns control to the scheduler, which may in turn call the
rebuilder to bring the dependent key up to date, and so on.

@@

The rebuilder receives three arguments: a \hs{key}, its current \hs{value}, and
the applicative \hs{task} that can be used to rebuild the \hs{key} if necessary.
The rebuilder first decides if the \hs{key} is \hs{dirty} by consulting
\hs{modTimes}: if the \hs{key} is not found, that must mean it has never been
built before; otherwise \hs{modTimeRebuilder} can see if any of the \hs{task}'s
dependencies (computed by \hs{dependencies}) are out of date. If the \hs{key} is
\hs{dirty}, we use \hs{run}~\hs{task} to rebuild it, and update the state with
the new modification time of the \hs{key}\footnote{The real \Make
relies on the file system to track file modification times, but we prefer
to make it explicit here.}; otherwise we can just return the current \hs{value}.

@@

The external system, which invokes \Excel's build engine, is required
to provide a transitively closed set of dirty bits. That is, if a
cell is changed, its dirty bit is set, as well as the dirty bit of any other
cell whose value might perhaps change as a result. It is OK to mark too many
cells as dirty; but not OK to mark too few.

@@

Fig.~\ref{fig-cloud-implementations} shows our models of several cloud build
systems. \Bazel, \CloudBuild and \Cloud \Shake are based on constructive traces
(\S\ref{sec-constructive-traces}), whereas \Buck and \Nix use their deep
variant (\S\ref{sec-deep-constructive-traces}).

@@

By switching to the \hs{topological} scheduler, we obtain a model of
Microsoft's \CloudBuild~-- an applicative build system that combines
conventional scheduling of statically known directed acyclic graphs of
dependencies with constructive traces~\cite{esfahani2016cloudbuild}. Note that
we need to convert a monadic \hs{ctRebuilder} into an applicative one by
applying an adapter function \hs{adaptRebuilder}, which unwraps a given
\hs{Task}~\hs{Applicative} and wraps it into \hs{Task}~\hs{Monad}.

### 2018-07-07 paper/6-engineering.tex
Final proofread

Our model assumes a world where the store is fully-defined, every \hs{k} is
associated with a \hs{v}, and every compute successfully completes returning a
valid value. In the real world, build systems frequently deal with errors, e.g.
``file not found'', or ``compilation failed''. We can model such failure
conditions by instantiating \hs{v} to either \hs{Maybe}~\hs{v} (for missing
values) or \hs{Either}~\hs{e}~\hs{v} (for exceptions of type \hs{e}). That can
model the values inside the store and the \hs{Task}, but because \hs{v} is
polymorphic for builds, it does not let the build system apply special behaviour
for errors, e.g. early aborting.

@@

\item The \hs{restarting} scheduler can be made parallel in a few ways, but the
most direct is to have $n$ threads reading entries from the list of keys. As
before, if a key requires something not yet built, it is added to the end -- the
difference is that sometimes things will be moved to the back of the queue not
because they are out of order but because of races with earlier nodes that had
not yet finished. As a consequence, over successive runs, potentially racey
dependencies will be separated, giving better parallelism over time.

\item The \hs{suspending} function can be made parallel by starting static
dependencies of a \hs{Task} in parallel, while dynamic dependencies are being
resolved, using the strategy described by~\citet{marlow2014haxl}.
\end{enumerate}

@@

Interestingly, there is significant interplay between all three sources of
impurity. Systems like \Bazel use various sandboxing techniques to guard against
missing dependencies, but none are likely to capture all dependencies right down
to CPU model and microcode version. Rules that do have untracked dependencies can be
marked as volatile, a technique \Excel takes with the \cmd{INDIRECT} function,
removing the untracked dependency at the cost of minimality.

@@

\item[Shallow builds] Building the end result, e.g. an installer package, often
involves many intermediate tasks. The intermediate results may be large, so some
cloud build systems are designed to build end products \emph{without
materialising} intermediate results, producing a so-called \emph{shallow build}
(see an example in~\S\ref{sec-background-bazel}). Some build systems can go even
further, integrating with the file system to only materialise the file when the
user accesses it \cite{gvfs}.
\end{description}

@@

\begin{figure}
\begin{subfigure}[b]{0.32\linewidth}
\centerline{\includegraphics[scale=0.28]{fig/frankenbuild-example-build.pdf}}
\vspace{-1.5mm}
\caption{Initial build}
\end{subfigure}
\begin{subfigure}[b]{0.32\linewidth}
\centerline{\includegraphics[scale=0.28]{fig/frankenbuild-example-clean.pdf}}
\vspace{-1.5mm}
\caption{Clean up and evict \cmd{main.prof}}
\end{subfigure}
\begin{subfigure}[b]{0.34\linewidth}
\centerline{\includegraphics[scale=0.28]{fig/frankenbuild-example-rebuild.pdf}}
\vspace{-1.5mm}
\caption{Build \cmd{main.prof} and \cmd{report.txt}}
\end{subfigure}
\vspace{-1.5mm}
\caption{A Frankenbuild example: (a)~build a human-readable profiling report for
\cmd{main.exe} from information dump \cmd{main.prof} produced by a profiling
tool, then record deep constructive traces in the cloud, (b)~remove built files
locally and evict \cmd{main.prof} from the cloud storage, (c)~build
\cmd{main.prof} by executing the profiler (profiling is non-deterministic, hence
the new hash), then build \cmd{report.txt} by downloading it from the matching
deep constructive trace in the cloud, resulting in a Frankenbuild because
\cmd{main.prof} and \cmd{report.txt} are inconsistent. New and evicted cloud
storage entries are highlighted; file hashes are shown in circles.
\label{fig-frankenbuild}}
\vspace{-4mm}
\end{figure}

@@

Some build systems, for example \Excel and \Ninja, are capable of recomputing a
task if either its dependencies change, \emph{or} the rule itself changes. For
example:

@@

\noindent
The build systems that have precise self-tracking are all ones which use a
\emph{non-embedded domain specific language} to describe build tasks. Those
which use a full programming language, e.g. \Shake, are faced with the challenge
of implementing equality on arbitrary task functions. For such build systems the
pessimistic assumption of saying that any change to the build system potentially
changes any build rule can often be used~--~the classic example being a makefile
depending on itself.

@@

Some computations are best described not by a chain of acyclic dependencies,
but by a loop. For example, \LaTeX~requires repeated rebuilding until it
reaches a fixed point, which can be directly expressed in build systems, such as
\Pluto~\cite{erdweg2015pluto}. Another example is \Excel, where a cell can
depend on itself, for example: \cmd{A1 = A1 + 1}. In such cases \Excel will
normally not execute anything, but if the ``Iterative Calculations'' feature is
enabled will execute the formula for a specified maximum number $N$ of times per
calculation (where $N$ is a setting that defaults to 100).

@@

A more compelling example is build tasks that produce multiple output
keys~--~for example, \cmd{ghc Foo.hs} produces both \cmd{Foo.hi} and \cmd{Foo.o}.
That can be represented by having a key whose value is a pair of file names, and
whose result is a pair of file contents. From that, the rule for \cmd{Foo.hi}
can be the first component of the result of the pair. Again, such an operation
can be encoded without polymorphic keys provided the pair of files (or a dummy
file representing the pair) is marked as changed if either of the contained
files change. Once again, polymorphic dependencies provide convenience rather
than power.

>>>>

Our model assumes a world where the store is fully-defined, every \hs{k} is
associated with a \hs{v}, and every compute successfully completes returning a
valid value. In the real world, build systems frequently deal with errors, e.g.
``file not found'' or ``compilation failed''. We can model such failure
conditions by instantiating \hs{v} to either \hs{Maybe}~\hs{v} (for missing
values) or \hs{Either}~\hs{e}~\hs{v} (for exceptions of type \hs{e}). That can
model the errors inside the store and the \hs{Task}, but because \hs{v} is
polymorphic for builds, it does not let the build system apply special behaviour
for errors, e.g. early aborting.

@@

\item The \hs{restarting} scheduler can be made parallel in a few ways, but the
most direct is to have $n$ threads reading keys from the build queue. As before,
if a key requires a dependency not yet built, it is moved to the end~--~the
difference is that sometimes keys will be moved to the back of the queue not
because they are out of date but because of races with earlier tasks that had
not yet finished. As a consequence, over successive runs, potentially racey
dependencies will be separated, giving better parallelism over time.

\item The \hs{suspending} scheduler can be made parallel by starting static
dependencies of a \hs{Task} in parallel, while dynamic dependencies are being
resolved, using the strategy described by~\citet{marlow2014haxl}.
\end{enumerate}

@@

Interestingly, there is significant interplay between all three sources of
impurity. Systems like \Bazel use various sandboxing techniques to guard against
missing dependencies, but none are likely to capture all dependencies right down
to the CPU model and microcode version. Tasks that do have untracked
dependencies can be marked as volatile, a technique \Excel takes with the
\cmd{INDIRECT} function, removing the untracked dependency at the cost of
minimality.

@@

\item[Shallow builds] Building the end result, e.g. an installer package, often
involves many intermediate tasks. The intermediate results may be large, so some
cloud build systems are designed to build end products \emph{without
materialising} intermediate results, producing a so-called \emph{shallow build}
(see an example in~\S\ref{sec-background-bazel}). Some build systems go even
further, integrating with the file system to only materialise the file when the
user accesses it \cite{gvfs}.
\end{description}

@@

\begin{figure}
\begin{subfigure}[b]{0.32\linewidth}
\centerline{\includegraphics[scale=0.28]{fig/frankenbuild-example-build.pdf}}
\vspace{-1.5mm}
\caption{Initial build}
\end{subfigure}
\begin{subfigure}[b]{0.32\linewidth}
\centerline{\includegraphics[scale=0.28]{fig/frankenbuild-example-clean.pdf}}
\vspace{-1.5mm}
\caption{Clean up and evict \cmd{main.prof}}
\end{subfigure}
\begin{subfigure}[b]{0.34\linewidth}
\centerline{\includegraphics[scale=0.28]{fig/frankenbuild-example-rebuild.pdf}}
\vspace{-1.5mm}
\caption{Build \cmd{main.prof} and \cmd{report.txt}}
\end{subfigure}
\vspace{-1.5mm}
\caption{A Frankenbuild example: (a)~build a human-readable profiling report for
\cmd{main.exe} from information dump \cmd{main.prof} produced by a profiling
tool, then record deep constructive traces in the cloud, (b)~remove built files
locally and evict \cmd{main.prof} from the cloud storage, (c)~build
\cmd{main.prof} by executing the profiler (profiling is non-deterministic, hence
the new hash value), then build \cmd{report.txt} by downloading it from the
matching deep constructive trace in the cloud, resulting in a Frankenbuild
because \cmd{main.prof} and \cmd{report.txt} are inconsistent. New and evicted
cloud storage entries are highlighted; file hashes are shown in circles.
\label{fig-frankenbuild}}
\vspace{-4mm}
\end{figure}

@@

Some build systems, for example \Excel and \Ninja, are capable of recomputing a
task if either its dependencies change, \emph{or} the task itself changes. For
example:

@@

\noindent
The build systems that have precise self-tracking are all ones which use a
\emph{non-embedded domain specific language} to describe build tasks. Those
which use a full programming language, e.g. \Shake, are faced with the challenge
of implementing equality on arbitrary task functions. For such build systems,
the pessimistic assumption that any change to the build system potentially
changes any build task can often be used~--~the classic example being a makefile
depending on itself.

@@

Some computations are best described not by a chain of acyclic dependencies,
but by a loop. For example, \LaTeX~requires repeated rebuilding until it
reaches a fixed point, which can be directly expressed in build systems, such as
\Pluto~\cite{erdweg2015pluto}. Another example is \Excel, where a cell can
depend on itself, for example: \cmd{A1 = A1 + 1}. In such cases \Excel will
normally not execute anything, but if the ``Iterative Calculations'' feature is
enabled \Excel will execute the formula for a specified maximum number $N$ of
times per calculation (where $N$ is a setting that defaults to 100).

@@

A more compelling example is build tasks that produce multiple output
keys~--~for example, \cmd{ghc Foo.hs} produces both \cmd{Foo.hi} and \cmd{Foo.o}.
That can be represented by having a key whose value is a pair of file names, and
whose result is a pair of file contents. From that, the task for \cmd{Foo.hi}
can be the first component of the result of the pair. Again, such an operation
can be encoded without polymorphic keys provided the pair of files (or a dummy
file representing the pair) is marked as changed if either of the contained
files change. Once again, polymorphic dependencies provide convenience rather
than power.

### 2018-07-07 paper/7-related.tex
Final proofread

Most build systems, when viewed at the level we talk, can be captured with minor
variations on the code presented in \S\ref{sec-implementations}. Below we list
some notable examples:

@@

\noindent
To achieve minimality, each separate command is traced at the OS-level, allowing
\Fabricate to record a trace entry stating that \cmd{gcc -c util.c} reads from
\cmd{util.c}. In future runs \Fabricate runs the script from start to finish,
skipping any commands where no inputs have changed. The key difference from our
\hs{Tasks} abstraction is that instead of supplying a mapping from output to action,
\Fabricate supplies a list of statements, in an order, without declaring what each line produces.
There is no need to schedule the statements, and not enough information to do so.

@@

While not typically considered build systems, self-adjusting computation is a
well studied area, and in particular the contrast between different formulations
has been thoroughly investigated, e.g. see~\cite{acar2007selfadjusting}.
Self-adjusting computations can automatically adjust to an external change
to their inputs. A classic example is a self-adjusting sorting algorithm, which
can efficiently (in $O(\log{n})$ time where $n$ is the length of the input)
recalculate the result given an incremental change of the input. While very
close to build systems in spirit, self-adjusting computations are mostly used
for in-memory computation and rely on the ability to dynamically allocate new
keys in the store for sharing intermediate computations~--~an intriguing feature
rarely seen in build systems (\Shake's oracles~\S\ref{sec-polymorphism} can be
used to model this feature to a limited degree).

A lot of research has been dedicated to finding efficient data structures and
algorithms for self-adjusting computations (with a few open-source
implementations, e.g. \Incremental~\cite{incremental}). We plan to investigate
how these insights can be utilised by build systems as future work.

>>>>

Most build systems, when viewed at the level we talk, can be captured with minor
variations of the code presented in \S\ref{sec-implementations}. Below we list
some notable examples:

@@

\noindent
To achieve minimality, each separate command is traced at the OS-level, allowing
\Fabricate to record a trace entry stating that \cmd{gcc -c util.c} reads from
\cmd{util.c}. In future runs \Fabricate runs the script from start to finish,
skipping any commands where no inputs have changed. The key difference from our
\hs{Tasks} abstraction is that instead of supplying a mapping from outputs to
tasks, \Fabricate supplies a list of statements, in an order, without declaring
what each line produces. There is no need to schedule the statements, and not
enough information to do so.

@@

While not typically considered build systems, self-adjusting computation is a
well studied area, and in particular the contrast between different formulations
has been thoroughly investigated, e.g. see~\citet{acar2007selfadjusting}.
Self-adjusting computations can automatically adjust to an external change
to their inputs. A classic example is a self-adjusting sorting algorithm, which
can efficiently (in $O(\log{n})$ time where $n$ is the length of the input)
recalculate the result given an incremental change of the input. While very
close to build systems in spirit, self-adjusting computations are mostly used
for in-memory computation and rely on the ability to dynamically allocate new
keys in the store for sharing intermediate computations~--~an intriguing feature
rarely seen in build systems (\Shake's oracles~\S\ref{sec-polymorphism} can be
used to model this feature to a limited degree).

A lot of research has been dedicated to finding efficient data structures and
algorithms for self-adjusting computations (with a few open-source
implementations, e.g. \Incremental by~\citet{incremental}). We plan to
investigate how these insights can be utilised by build systems as future work.
